A White Noise Approach to Evolutionary Ecology

### 2020-04-10 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
improved formatting

where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x,y)\nu(y,t)dy$ for some non-negative and bounded bivariate function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. An example of a nonlocal effect is the competition for a common resource among all individuals regardless of their trait value so that individuals with differing trait values effect each others fitness. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. Now that these technicalities have been treated we can model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

@@

and make use of the convention $f=g$ if $\|f-g\|_t=0$ for all $t\geq0$.

@@

Unfortunately, just as with cream cheese spread across too much toast, the blob perspective of the population may exhibit some complicated spatial discontinuities. However, for spatial dimension $d=1$, it turns out that $\mathscr{X}_t$ is absolutely continuous with respect to the Lebesgue measure for each $t\geq0$ [@Konno1988; @Reimers1989]. This means that we can write $\mathscr{X}_t(D)=\int_D \nu(x,t)dx$ for some density process $\nu(x,t)$. Setting $\lambda=1$ and $m=\ln\mathscr{W}$ this density process satisfies the SPDE
\begin{equation}\label{neutral_SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Since $\nu(x,t)$ is not generally differentiable in $x$ or $t$, the derivatives in expression (\ref{neutral_SPDE}) are taken in the _weak_ sense [sensu @lawrenceevans2010]. That is, to rigorously interpret SPDE (\ref{neutral_SPDE}), we integrate the solution $\nu(x,t)$ against functions $f\in C_b^2(\mathbb{R})$, where $C_b^2(\mathbb{R})$ is the set of bounded and twice continuously differentiable functions on $\mathbb{R}$. Hence, equation (\ref{neutral_SPDE}) is just an abbreviation for

@@

Recall, we use $\nu(x,t)$ to denote the density of a superprocess, given it exists. Assuming the interactions manifest only in the reproductive law and that spatial movement follows Brownian motion with diffusion parameter $\sqrt\mu$ independent of both $X_t$ and $x$, Li [-@zeng1998absolute] proved a result that implies the interacting superprocess on one dimensional trait space has a density $\nu(x,t)$ which is non-negative, integrable, continuous in time and space and satisfies the SPDE
\begin{equation}\label{SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Comparing equation (\ref{SPDE}) to equation (3.5) of Li [-@zeng1998absolute], our $m$ and $V$ correspond to Li's $b$ and $c$ respectively. It is important to note that, under the assumptions made in Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] and Li [-@zeng1998absolute], equation (\ref{SPDE}) is only formal when $m(y,x)$ is bounded across all combinations of $y\geq0$ and $x\in\mathbb{R}$. However, recalling our condition $m(y,x)\leq r\in\mathbb{R}$, the growth rates we consider are only bounded above. Yet, in the proof of the construction of the interacting superprocess as the limit of rescaled branching diffusions, Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] assumed $m(y,x)$ to be bounded to guarantee the total mass process will have finite mean and variance, for finite $t\geq0$. This allowed the authors to employ a tightness criterion for sequences of measures and show the rescaled processes converge to a superprocess with finite total mass. Li's [-@zeng1998absolute] result builds directly off of Méléard and Roelly's construction, inheriting the assumption of boundedness for $m(y,x)$. However, in Li [-@zeng1998absolute], the sufficiency of $m(y,x)$ being bounded above is even more clear since Li works explicitly with a common upperbound for both $m(y,x)$ and $V$. Hence, one can repeat the necessary proofs replacing the assumption that $m(y,x)$ is bounded with the assumption that $m(y,x)$ is merely bounded above to derive the same results.

@@

In SM \S\ref{SDE_DERIV} we show SDE for $N(t)$, $\bar x(t)$ and $\sigma^2(t)$ can be expressed as
\begin{subequations}\label{gen_eqns}
\begin{equation}\label{N}
dN(t)=\bar m(t)N(t)dt+\sqrt{V N(t)}dW_1(t),
\end{equation}
\begin{equation}\label{xbar_gen}
d\bar x(t)=\mathrm{Cov}_t\Big(x,m(\nu,x)\Big)dt+\sqrt{V\frac{\sigma^2(t)}{N(t)}}dW_2(t),
\end{equation}
\begin{equation}\label{sig2_gen}
d\sigma^2(t)=\left(\mathrm{Cov}_t\Big((x-\bar x(t))^2,m(\nu,x)\Big)+\mu-V\frac{\sigma^2(t)}{N(t)}\right)dt+\sqrt{V\frac{\overline{(x-\bar x(t))^4}-\sigma^4(t)}{N(t)}}dW_3(t),
\end{equation}
\end{subequations}
where $W_1$, $W_2$ and $W_3$ are standard Brownian motions. We note that conditions on the growth rate $m$ to guarantee existence and uniqueness of solutions to (\ref{xbar_gen}) and (\ref{sig2_gen}) have yet to be investigated. However, our results on the deterministic PDE suggest that $m(y,x)$ bounded above and differentiable in both arguments is sufficient. Dividing by $dt$ one can interpret equations (\ref{gen_eqns}) as if they are ordinary differential equations, but this not technically rigorous since Brownian motion is nowhere differentiable with respect to time. In SM \S\ref{SDE_DERIV} we show that in general $W_1$ is independent of both $W_2$ and $W_3$, but $W_2$ and $W_3$ covary.

@@

By assuming $\nu(x,t)$ can be approximated by a Gaussian curve for each $t\geq0$, expressions (\ref{N}), (\ref{xbar_gen}) and (\ref{xbar_gen}) transform into efficient tools for deriving the dynamics of populations given a fitness function $m(\nu,x)$. Gaussian phenotypic distributions are often obtained through Gaussian, exponential or weak selection approximations together with a simplified model of inheritance and random mating [@pmid17248993; @Turelli1984; @Turelli1986; @Turelli2017; @9780471986539]. Alternatively, it has been shown that a Gaussian distribution can provide a reasonable approximation even when selection is strong and non-Gaussian [@pmid7851785]. However, our approach adds an additional layer of difficulty. Even with Gaussian selection, the resulting solution to SPDE (\ref{SPDE}) will only be a Gaussian curve in expectation, assuming a Gaussian initial condition. Yet this difficulty is not as challenging as it may first appear. Indeed, since SPDE (\ref{SPDE}) can be derived as a diffusion limit we know that, under the appropriate assumptions on selection, genetic architecture and reproduction, the stochastic departure from a Gaussian curve is negligible when the ratio $V/N$ is small (i.e., when the variance in reproductive output is much smaller than the population size). In SM \S\ref{numerical} we demonstrate this result using numerical methods. Mathematically, this requirement restricts model parameters to regions that maintain large population sizes. Biologically, this implies populations are not at risk of extinction. Hence, models developed in this framework are not suitable for studying colonization-extinction dynamics or evolutionary rescue. Allowing for these restrictions, we may safely assume that $\nu$ is approximately Gaussian and justify writing
\begin{equation}
\nu(x,t)=\frac{N(t)}{\sqrt{2\pi\sigma^2(t)}}\exp\left(-\frac{\big(x-\bar x(t)\big)^2}{2\sigma^2(t)}\right).
\end{equation}
Under this assumption we find in SM \S\ref{cov2deriv} the results (suppressing the dependency on $t$)
\begin{equation}\label{covxm}
\mathrm{Cov}(x,m)=\sigma^2\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right),
\end{equation}
\begin{equation}
\mathrm{Cov}\Big((x-\bar x)^2,m\Big)=2\sigma^4\left(\frac{\partial\bar m}{\partial\sigma^2}-\overline{\frac{\partial m}{\partial\sigma^2}}\right)
\end{equation}
and $\overline{(x-\bar x)^4}=3\sigma^4$. Equation (\ref{covxm}) is the continuous time equivalent to equation (9) in Lande [-@Lande1976]. In particular, these results imply
\begin{subequations}\label{no_inher}
\begin{equation}\label{xbar}
d\bar x=\sigma^2\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right)dt+\sqrt{V\frac{\sigma^2}{N}}dW_2,
\end{equation}
\begin{equation}\label{G}
d\sigma^2=2\sigma^4\left(\frac{\partial\bar m}{\partial\sigma^2}-\overline{\frac{\partial m}{\partial\sigma^2}}\right)dt +\left(\mu-V\frac{\sigma^2}{N}\right)dt+\sigma^2\sqrt{\frac{2V}{N}}dW_3.
\end{equation}
\end{subequations}
These equations allow us to derive the response in trait mean and variance by taking derivatives of fitness, a much more straightforward operation than calculating a covariance for general phenotypic distributions. Note that in the above expressions, the partial derivatives of $\bar m$ represent frequency independent selection and the averaged partial derivatives of $m$ represent frequency dependent selection. This relationship has already been pointed out by Lande [-@Lande1976] for the evolution of the mean trait, but here we see an analogous relationship holds also for the evolution of trait variance.

@@

In this section we demonstrate the use of our framework by developing a model of diffuse coevolution across a guild of $S$ species whose interactions are mediated by resource competition along a single niche axis. Because our approach treats abundance dynamics and evolutionary dynamics simultaneously, this model allows us to investigate the relationship between selection gradients and competition coefficients.

@@

```{r, message=FALSE, echo=F, fig.align='center', fig.cap='\\label{temporal}Temporal dynamics of mean trait (top), additive genetic variance (middle) and abundance (bottom) for the scenario of weak competition (left) and strong competition (right). Red lines indicate average trend across species.'}
knitr::include_graphics('community_dynamics.jpg')
```

>>>>

where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x,y)\nu(y,t)dy$ for some non-negative and bounded bivariate function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. An example of a nonlocal effect comes from competition for a common resource among all individuals in a population regardless of their trait value. In this case all individuals in the population effect each others fitness' even though they may exhibit diverse trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. Now that these technicalities have been treated we can model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE

@@

and make use of the convention $f=g$ if $\|f-g\|_t=0$ for all $t\geq0$. Later, when we investigate stochastic dynamics of the abundance density $\nu(x,t)$, we will find $\sqrt\nu\in\mathscr{N}_2$. This enables us to utilize the heuristics developed in this section for the derivation of SDE describing the stochastic dynamics of $N(t)$, $\bar x(t)$ and $\sigma^2(t)$.

@@

Unfortunately, just as with cream cheese spread across too much toast, the blob perspective of the population may exhibit some complicated spatial discontinuities. However, for spatial dimension $d=1$, it turns out that $\mathscr{X}_t$ is absolutely continuous with respect to the Lebesgue measure for each $t\geq0$ [@Konno1988; @Reimers1989]. This means that we can write $\mathscr{X}_t(D)=\int_D \nu(x,t)dx$ for some density process $\nu(x,t)$. Setting $\lambda=1$ and $m=\ln\mathscr{W}$ this density process satisfies the SPDE
\begin{equation}\label{neutral_SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Since $\nu(x,t)$ is not generally differentiable in $x$ or $t$, the derivatives in expression (\ref{neutral_SPDE}) are taken in the _weak_ sense [for examples, see @lawrenceevans2010]. That is, to rigorously interpret SPDE (\ref{neutral_SPDE}), we integrate the solution $\nu(x,t)$ against functions $f\in C_b^2(\mathbb{R})$, where $C_b^2(\mathbb{R})$ is the set of bounded and twice continuously differentiable functions on $\mathbb{R}$. Hence, equation (\ref{neutral_SPDE}) is just an abbreviation for

@@

Recall, we use $\nu(x,t)$ to denote the density of a superprocess, given it exists. Assuming the interactions manifest only in the reproductive law and that spatial movement follows Brownian motion with diffusion parameter $\sqrt\mu$ independent of both $X_t$ and $x$, Li [-@zeng1998absolute] proved a result that implies the interacting superprocess on one dimensional trait space has a density $\nu(x,t)$ which is non-negative, integrable, continuous in time and space and satisfies the SPDE
\begin{equation}\label{SPDE}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)+\sqrt{V\nu(x,t)}\dot W(x,t).
\end{equation}
Comparing equation (\ref{SPDE}) to equation (3.5) of Li [-@zeng1998absolute], our $m$ and $V$ correspond to Li's $b$ and $c$ respectively. It is important to note that, under the assumptions made in Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] and Li [-@zeng1998absolute], equation (\ref{SPDE}) is only formal when $m(h,x)$ is bounded across all combinations of $h\geq0$ and $x\in\mathbb{R}$. However, recalling our condition $m(h,x)\leq r\in\mathbb{R}$, the growth rates we consider are only bounded above. Yet, in the proof of the construction of the interacting superprocess as the limit of rescaled branching diffusions, Méléard and Roelly [-@meleard1992interacting; -@Mlard1993] assumed $m(h,x)$ to be bounded to guarantee the total mass process will have finite mean and variance, for finite $t\geq0$. This allowed the authors to employ a tightness criterion for sequences of measures and show the rescaled processes converge to a superprocess with finite total mass. Li's [-@zeng1998absolute] result builds directly off of Méléard and Roelly's construction, inheriting the assumption of boundedness for $m(h,x)$. However, in Li [-@zeng1998absolute], the sufficiency of $m(h,x)$ being bounded above is even more clear since Li works explicitly with a common upperbound for both $m(h,x)$ and $V$. Hence, one can repeat the necessary proofs replacing the assumption that $m(h,x)$ is bounded with the assumption that $m(h,x)$ is merely bounded above to derive the same results.

@@

In SM \S\ref{SDE_DERIV} we show SDE for $N(t)$, $\bar x(t)$ and $\sigma^2(t)$ can be expressed as
\begin{subequations}\label{gen_eqns}
\begin{equation}\label{N}
dN(t)=\bar m(t)N(t)dt+\sqrt{V N(t)}dW_1(t),
\end{equation}
\begin{equation}\label{xbar_gen}
d\bar x(t)=\mathrm{Cov}_t\Big(x,m(\nu,x)\Big)dt+\sqrt{V\frac{\sigma^2(t)}{N(t)}}dW_2(t),
\end{equation}
\begin{equation}\label{sig2_gen}
d\sigma^2(t)=\mathrm{Cov}_t\Big((x-\bar x(t))^2,m(\nu,x)\Big)dt+\left(\mu-V\frac{\sigma^2(t)}{N(t)}\right)dt+\sqrt{V\frac{\overline{(x-\bar x(t))^4}-\sigma^4(t)}{N(t)}}dW_3(t),
\end{equation}
\end{subequations}
where $W_1$, $W_2$ and $W_3$ are standard Brownian motions. We note that conditions on the growth rate $m$ to guarantee existence and uniqueness of solutions to (\ref{xbar_gen}) and (\ref{sig2_gen}) have yet to be investigated. However, our results on the deterministic PDE suggest that $m(y,x)$ bounded above and differentiable in both arguments is sufficient. Dividing by $dt$ one can interpret equations (\ref{gen_eqns}) as if they are ordinary differential equations, but this not technically rigorous since Brownian motion is nowhere differentiable with respect to time. In SM \S\ref{SDE_DERIV} we show that in general $W_1$ is independent of both $W_2$ and $W_3$, but $W_2$ and $W_3$ covary.

@@

By assuming $\nu(x,t)$ can be approximated by a Gaussian curve for each $t\geq0$, expressions (\ref{N}), (\ref{xbar_gen}) and (\ref{sig2_gen}) transform into efficient tools for deriving the dynamics of populations given a fitness function $m(\nu,x)$. Gaussian phenotypic distributions are often obtained through Gaussian, exponential or weak selection approximations together with a simplified model of inheritance and random mating [@pmid17248993; @Turelli1984; @Turelli1986; @Turelli2017; @9780471986539]. Alternatively, it has been shown that a Gaussian distribution can provide a reasonable approximation even when selection is strong and non-Gaussian [@pmid7851785]. However, our approach adds an additional layer of difficulty. Even with Gaussian selection, the resulting solution to SPDE (\ref{SPDE}) will only be a Gaussian curve in expectation, assuming a Gaussian curve as the initial condition. Yet this difficulty is not as challenging as it may first appear. Indeed, since SPDE (\ref{SPDE}) can be derived as a diffusion limit we know that, under the appropriate assumptions on selection, genetic architecture and reproduction, the stochastic departure from a Gaussian curve is negligible when the ratio $V/N$ is small (i.e., when the variance in reproductive output is much smaller than the population size). In SM \S\ref{numerical} we demonstrate this result using numerical methods. Mathematically, this requirement restricts model parameters to regions that maintain large population sizes. Biologically, this implies populations are not at risk of extinction. Hence, models developed in this framework are not suitable for studying colonization-extinction dynamics or evolutionary rescue. Allowing for these restrictions, we may safely assume that $\nu$ is approximately Gaussian and justify writing
\begin{equation}
\nu(x,t)=\frac{N(t)}{\sqrt{2\pi\sigma^2(t)}}\exp\left(-\frac{\big(x-\bar x(t)\big)^2}{2\sigma^2(t)}\right).
\end{equation}
Under this assumption we find in SM \S\ref{cov2deriv} the results (suppressing the dependency on $t$)
\begin{equation}\label{covxm}
\mathrm{Cov}(x,m)=\sigma^2\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right),
\end{equation}
\begin{equation}
\mathrm{Cov}\Big((x-\bar x)^2,m\Big)=2\sigma^4\left(\frac{\partial\bar m}{\partial\sigma^2}-\overline{\frac{\partial m}{\partial\sigma^2}}\right)
\end{equation}
and $\overline{(x-\bar x)^4}=3\sigma^4$. Equation (\ref{covxm}) is the continuous time equivalent to equation (9) in Lande [-@Lande1976]. In particular, these results imply
\begin{subequations}\label{no_inher}
\begin{equation}\label{xbar}
d\bar x=\sigma^2\left(\frac{\partial\bar m}{\partial\bar x}-\overline{\frac{\partial m}{\partial\bar x}}\right)dt+\sqrt{V\frac{\sigma^2}{N}}dW_2,
\end{equation}
\begin{equation}\label{G}
d\sigma^2=2\sigma^4\left(\frac{\partial\bar m}{\partial\sigma^2}-\overline{\frac{\partial m}{\partial\sigma^2}}\right)dt +\left(\mu-V\frac{\sigma^2}{N}\right)dt+\sigma^2\sqrt{\frac{2V}{N}}dW_3.
\end{equation}
\end{subequations}
These equations allow us to derive the response in trait mean and variance by taking derivatives of fitness, a much more straightforward operation than calculating a covariance for general phenotypic distributions. Note that in the above expressions, the partial derivatives of $\bar m$ represent frequency independent selection and the averaged partial derivatives of $m$ represent frequency dependent selection. This relationship has already been pointed out by Lande [-@Lande1976] for the evolution of the mean trait, but here we see an analogous relationship holds also for the evolution of trait variance.

@@

In this section we demonstrate the use of our framework by developing a model of diffuse coevolution across a guild of $S$ species whose interactions are mediated by resource competition along a single niche axis. Because our approach treats abundance dynamics and evolutionary dynamics simultaneously, this model allows us to investigate the relationship between selection gradients and competition coefficients, which we carry out in what follows.

@@

```{r, message=FALSE, echo=F, fig.align='center', out.width='100%', fig.cap='\\label{temporal}Temporal dynamics of mean trait (top), additive genetic variance (middle) and abundance (bottom) for the scenario of weak competition (left) and strong competition (right). Red lines indicate average trend across species.'}
knitr::include_graphics('community_dynamics.jpg')
```
